---
title: "Dynamic Models for Poisson and Binomial Time Series"
author: "Gregor Zens"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Dynamic Models for Poisson and Binomial Time Series}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 7,
  fig.height = 4.2
)
set.seed(1)
library(DynCount)
# Short MCMC runs keep the vignette fast to build; use longer runs in practice.
NSAVE <- 1000L
NBURN <- 1000L
```

## Introduction

`DynCount` fits state-space models to non-Gaussian time series. A
latent trajectory \(z_t\) evolves with one of two dynamics,
\[
  z_t = \mu + \rho\, z_{t-1} + \varepsilon_t,
\]
a first-order random walk (`latent_dynamics = "rw"`, i.e. \(\rho = 1\)) or a
stationary AR(1) process (`latent_dynamics = "ar1"`, with \(\rho\) estimated and
constrained to \((-1, 1)\)). The scalar \(\mu\) is zero unless a drift/intercept
is included. The observations are linked to it through one of two observation
models:

* **Poisson**, with a log link: \(y_t \sim \mathrm{Poisson}(e^{z_t})\);
* **Binomial**, with a logit link: \(y_t \sim \mathrm{Binomial}(m_t,\,
  \mathrm{logit}^{-1}(z_t))\), where the trials \(m_t\) are known.

An optional known `offset` \(o_t\) may be added to the linear predictor of either
observation model -- a log-exposure for the Poisson mean, \(e^{o_t + z_t}\), or a
shift of the binomial logit. It is a fixed, user-supplied input, not part of the
latent process \(z_t\), and defaults to zero.

The model is estimated by MCMC. The distribution of the increments
\(\varepsilon_t = z_t - \mu - \rho z_{t-1}\) is controlled by the `innovations`
argument and can be Gaussian, Student-t, a finite scale mixture of normals, or a
stochastic volatility process. For either family, zeros can optionally be
handled by time-constant zero inflation.

The package implements and extends the methodology of Zens and Bijak
(2026), *The Annals of Applied Statistics*,
[doi:10.1214/26-AOAS2171](https://doi.org/10.1214/26-AOAS2171).

Note: the MCMC runs below use short chains (`nsave` = `r NSAVE`,
`nburn` = `r NBURN`) and, for the two shipped series, a shortened window,
so that the vignette builds quickly. For real analyses use longer chains and
the full series.

## Simulating data

The simulation helpers generate data with a known latent path,
which is useful for checking recovery.

```{r simulate}
sim <- simulate_dynamic_poisson(n = 80, sigma = 0.18, log_rate0 = 2.5, seed = 1)
str(sim, max.level = 1)
plot(sim$y, type = "h", xlab = "time", ylab = "count",
     main = "Simulated Poisson random walk")
lines(sim$rate, col = "steelblue", lwd = 2)
```

## Fitting a Poisson model (no zero inflation)

The main entry point is `fit_dynamic_model()`. The defaults give an ordinary
Poisson random walk with Gaussian increments.

```{r fit-poisson}
fit <- fit_dynamic_model(sim$y, family = "poisson",
                         nsave = NSAVE, nburn = NBURN, seed = 1)
fit
summary(fit)
```

`plot_fitted()` overlays the posterior of the fitted mean on the data, and
`plot_latent()` shows the latent log-rate trajectory with a credible band.

```{r plot-fitted}
plot_fitted(fit)
```

```{r plot-latent}
plot_latent(fit)
```

## Forecasting

Forecasts are produced **during** sampling. Request the horizon when fitting
with `horizon = H`: `H` extra latent states are appended to the state vector
and sampled as missing values inside the MCMC, and observations are drawn from
the observation model at the sampled forecast states, so the intervals
propagate parameter, state and innovation uncertainty. `forecast()` then
extracts and summarises the stored draws.

```{r forecast}
fit_fc <- fit_dynamic_model(sim$y, family = "poisson", horizon = 8,
                            nsave = NSAVE, nburn = NBURN, seed = 1)
fc <- forecast(fit_fc)
fc                                 # prints the forecast path
fc$final                           # the single 8-step-ahead forecast
plot_forecast(fit_fc)
```

The object stores the full forecast path (`fc$summary`, one row per horizon)
and, separately, the final h-step-ahead prediction (`fc$final`,
`fc$final_draws`). For long horizons, or with the heavier-tailed innovation
structures (`"t"`, `"mixture"`, `"sv"`), use longer chains than the short ones
shown here and, as with any MCMC output, check convergence before relying on
the results.

## AR(1) latent dynamics

Setting `latent_dynamics = "ar1"` estimates an autoregressive coefficient
\(\rho\) instead of fixing it at 1, jointly with an intercept \(\mu\). The pair
is drawn by an exact conjugate Gibbs step, with \(\rho\) **truncated to the
stationary region** \((-1, 1)\), so the posterior places mass only on
stationary processes. The
random walk is the special case \(\rho = 1\). **AR(1) always includes an
intercept**: `include_mu` is enabled automatically, giving the process a
non-zero stationary mean \(\mu / (1 - \rho)\).

```{r ar1}
# a genuinely stationary AR(1) log-rate: stationary mean 4, so mu = 4 * (1 - rho)
sim_ar <- simulate_dynamic_poisson(150, sigma = 0.2, log_rate0 = 4,
                                   rho = 0.9, mu = 0.4, seed = 3)
# no need to set include_mu: AR(1) enables the intercept automatically
fit_ar <- fit_dynamic_model(sim_ar$y, latent_dynamics = "ar1",
                            nsave = NSAVE, nburn = NBURN, seed = 3)
summary(fit_ar)                    # reports the posteriors of ar1_rho and mu
```

## Offset

For Poisson data with varying exposure, pass a known `offset` (a log-exposure
term); the mean becomes \(\exp(\text{offset}_t + z_t)\). Supply `forecast_offset`
at fitting time for the future exposures.

```{r offset}
expo  <- log(runif(120, 50, 200))  # known exposure, e.g. population at risk
sim_o <- simulate_dynamic_poisson(120, sigma = 0.12, log_rate0 = -3.5,
                                  offset = expo, seed = 4)
fit_o <- fit_dynamic_model(sim_o$y, offset = expo, horizon = 6,
                           forecast_offset = log(120),
                           nsave = NSAVE, nburn = NBURN, seed = 4)
forecast(fit_o)$final
```

## A robust (Student-t) fit on the example data

The package ships two real weekly count series of irregular maritime crossings.
`med_weekly` (Mediterranean route) has larger counts and few zeros; the
Student-t innovation makes the latent path robust to the occasional large
week-to-week jump. For a fast build we demonstrate this using a recent window of the series.

```{r med-t}
data(med_weekly)
med <- tail(med_weekly$count, 120)
fit_med <- fit_dynamic_model(med, family = "poisson",
                             innovations = "t",
                             nsave = NSAVE, nburn = NBURN, seed = 2)
summary(fit_med)
```

The posterior of the degrees-of-freedom parameter `t_df` indicates how heavy
the increment tails are, with smaller values indicating heavier tails.

## Zero inflation and structural zeros

`uk_weekly` (English Channel crossings) has many zeros. Turning on zero
inflation lets the model separate *structural* zeros (a time-constant gate that
switches the count off) from *sampling* zeros (zeros the Poisson process
produces on its own). We use the zero-heavy early window of the series here.

```{r fit-zip}
data(uk_weekly)
uk <- uk_weekly$count[1:130]
mean(uk == 0)                         # many zeros
fit_zip <- fit_dynamic_model(uk, family = "poisson",
                             zero_inflation = TRUE,
                             nsave = NSAVE, nburn = NBURN, seed = 3)
summary(fit_zip)
```

`structural_zero_prob()` reports, for each observed zero, the posterior
probability that it is structural. 

```{r structural}
sz <- structural_zero_prob(fit_zip, zeros_only = TRUE)
head(sz, 10)
plot_zero_inflation(fit_zip)
```

Interpreting the table: `p_structural` close to 1 flags a zero that the latent
rate cannot easily explain (e.g., the underlying rate was high, so a Poisson zero
would be unlikely), whereas `p_structural` near 0 marks a zero that is
consistent with a genuinely low rate.

### Conditional versus unconditional fits and replicates

Under zero inflation the observed count is \(y_t = v_t \tilde y_t\), where the
gate \(v_t \sim \mathrm{Bernoulli}(\pi_{\text{open}})\) switches the count off
and \(\tilde y_t\) comes from the Poisson/binomial observation model. The fit stores **both** flavours of in-sample quantities:

* **Unconditional** (`fitted`, `yrep`; the defaults returned by
  `predict()`): the gate is included, so the replicates reproduce the
  structural zeros. Use these for **posterior predictive checks**.
* **Conditional on the gate being open** (`fitted_open`, `yrep_open`;
  `predict(fit, conditional = TRUE)`): the latent-implied mean and a replicate
  drawn straight from the observation model, describing the latent intensity
  process.

For models without zero inflation the two versions are identical. Response
forecasts (`forecast()`, `forecast_y`) are always unconditional -- the gate is
applied to each forecast draw.

```{r ppc-zip}
# zero proportion in the data vs both flavours of replicate
c(observed      = mean(uk == 0),
  yrep          = mean(fit_zip$draws$yrep == 0),        # gate applied: comparable
  yrep_open     = mean(fit_zip$draws$yrep_open == 0))   # gate-open only: too few zeros
```

## A binomial model with known trials

The binomial branch keeps the same interface; the only addition is the known
number of `trials`.

```{r binomial}
simb <- simulate_dynamic_binomial(n = 80, sigma = 0.12, trials = 50, seed = 4)
fit_bin <- fit_dynamic_model(simb$y, family = "binomial", trials = simb$trials,
                             horizon = 8, forecast_trials = 50,
                             nsave = NSAVE, nburn = NBURN, seed = 4)
summary(fit_bin)
plot_fitted(fit_bin)
```

Forecasting works the same way; supply `forecast_trials` (at fitting time) for
the future trial sizes.

```{r binomial-forecast}
fc_bin <- forecast(fit_bin)
fc_bin$summary
```

Zero inflation is available for the binomial family too: a structural-zero
gate sits in front of the `Binomial(m, p)` process, exactly as for the
Poisson. Set `zero_inflation = TRUE` (or `zeros = "inflated"`) and read the
per-zero diagnostics with `structural_zero_prob()`.

```{r binomial-zip}
simz <- simulate_dynamic_binomial(n = 80, sigma = 0.1, trials = 40, logit0 = 1.5,
                                  zero_inflation = 0.2, seed = 7)
fit_bz <- fit_dynamic_model(simz$y, family = "binomial", trials = 40,
                            zero_inflation = TRUE,
                            nsave = NSAVE, nburn = NBURN, seed = 7)
head(structural_zero_prob(fit_bz))
```

## Choosing and changing priors

Every prior hyperparameter is exposed through `dynamic_prior()`; printing the
object shows the current settings.

```{r priors}
dynamic_prior()
# a tighter prior favouring smoother latent paths
pr <- dynamic_prior(var_shape = 10, var_rate = 0.2)
fit_smooth <- fit_dynamic_model(sim$y, prior = pr,
                                nsave = NSAVE, nburn = NBURN, seed = 1)
```

## References

Zens, G. and Bijak, J. (2026). Dynamic Count Models with Flexible Innovation
Processes for Irregular Maritime Migration. *The Annals of Applied Statistics*,
20(2), 1671--1690.
[doi:10.1214/26-AOAS2171](https://doi.org/10.1214/26-AOAS2171)
